---
title: "tidymetadata"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{tidymetadata tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(list(echo=TRUE,eval=TRUE,cache=FALSE,warning=FALSE,message=FALSE, fig.height=3, fig.width=5))
```


# Using tidymetadata-package with Luxembourdg Income Study data

Install the package with

```{r, eval=FALSE}
devtools::install_github("muuankarski/tidymetadata")
```

Tidymetadata-package provides a technical backbone for *tidy metadata* logic to work with social survey data using tidyverse tools.

If not familiar with tidy data or tidyverse, please familiarise yourself with:

- [Wickham, Hadley. 2014. ‘Tidy Data’. Journal of Statistical Software 59 (10). doi:10.18637/jss.v059.i10.](http://vita.had.co.nz/papers/tidy-data.html)
- [R for Data Science: Tidy Data](http://r4ds.had.co.nz/tidy-data.html)
- <http://tidyverse.org/>

Tidymetadata is not replacement for dedicated packages to work with survey data in general such as [`survey`](http://r-survey.r-forge.r-project.org/survey/), but it complements them by providing a framework to deal with the *labelled* structure typical for social survey data. Such structure is not optimal to work with in  R and tidyverse. Basically, tidymetadata provides functions to **separate** the **data** and **metadata** from *labelled* data.frame imported in R with [`haven`](http://haven.tidyverse.org/) and how to put it back together. This approach comes especially handy when building shiny-applications on top of such data. Try first running the shiny-app wrapped with the package with `tidymetadata::runExample()`.

## Download SPSS, Stata and SAS training data from Luxembourg Income Study

Survey datas are in most cases still disseminated in either SAS, SPSS or Stata formats with metadata included in variable and value labels. We first download the **demo** files freely available at from [Luxembourg Income Study](http://www.lisdatacenter.org/).


```{r}
library(tidyverse)
if (!file.exists("./datasets/data.sav")){

  dir.create("./datasets")
  # SPSS-file
  download.file("http://www.lisdatacenter.org/wp-content/uploads/it04ip.sav", "./datasets/data.sav")
  # Stata-file
  download.file("http://www.lisdatacenter.org/wp-content/uploads/it04ip.dta", "./datasets/data.dta")
  # SAS-file
  download.file("http://www.lisdatacenter.org/wp-content/uploads/it04ip.sas7bdat", "./datasets/data.sas7bdat")

}

```

Package [`haven`](http://haven.tidyverse.org/) provides handy functions for reading in such formats


```{r}
library(haven)
spss  <- read_sav("./datasets/data.sav")
stata <- read_dta("./datasets/data.dta")
sas   <- read_sas("./datasets/data.sas7bdat")
```

They all have the same number of cases and variables, and seem to share similar structure with variable labels and value labels as attributes. However, when applying `str()` function for all three datas we can spot that imported spss and stata files seem to have identical structure and somewhat richer than sas file has.

## Create "tidy metadata"

Once we have datafiles in R, we use function `tidymetadata::create_meta()` to create tidy metadata. `tidymetadata::create_meta()` extracts the metadata information from variable attributes and creates a tidy data frame with columns:

- `code` for variable code
- `name` for variable name
- `value` for each value for each labelled variable
- `label` for each value label for each labelled variable
- `class` for wheter the variable in question is `factor`, `numeric` or `character`

**As for rows**, tidy metadata has a single row for a unique `numeric` or `character` type of variable. Such variables are not labelled ie. have no additional information within the object. As for `factor` variables, tidy metadata has **one row per one value ie. factor level**. For instance, two class *gender* variable would have two rows in tidy metadata, one for both values.

Lets apply the `tidymetadata::create_meta()`-function for each data:

```{r}
library(tidymetadata)
meta_spss <- create_metadata(spss)
meta_stata <- create_metadata(stata)
meta_sas <- create_metadata(sas)
```

And, lets subset each data into variable `sex` and pile up the resulting rows:

```{r}
meta_spss$type ="spss"
meta_stata$type ="stata"
meta_sas$type ="sas"

bind_rows(
          meta_spss[meta_spss$code == "sex",],
          meta_stata[meta_stata$code == "sex",],
          meta_sas[meta_sas$code == "sex",]
          ) %>% 
  knitr::kable()

```

We can see that SPSS and Stata files produce identical metadatas for variable `sex`, but SAS file is lacking the value label information. For this reason we should work only using either Stata or SPSS files in this case, and for this particular vignette we will choose the SPSS file. 

## Stripping data from attributes

Once we have extracted the metadata information from the labelled data, we can strip off this information and leave only the bare data with `numeric/character` content in the variables using `tidymetadata::strip_attributes()` function with no attributes whatsoever.

Before stripping the attributes we get:

```{r}
str(spss[,1:10])
```

And after removing attributes the output for same columns is:

```{r}
# remove_attributes
data_spss <- tidymetadata::strip_attributes(spss)
str(data_spss[,1:10])
```

So, now we have two tidy datas, `data_spss` and `meta_spss`, and we can use whatever methods we need to for  analysing `data_spss`. 

## Labelling the data

For labelling the data with the `meta_spss` we can use function `tidymwetadata::label_data()` that basically uses `match()` for mapping the numeric values with corresponding value labels. Below we are labeling the variable `sex`.

```{r}
data_spss$sex_lab <- tidymetadata::label_data(data_spss, 
                                               variable.data ="sex", 
                                               variable.meta = "sex",
                                               metadata=meta_spss, 
                                               into.factor=FALSE)

head(data_spss[c("sex","sex_lab")])
data_spss$sex_lab <- NULL # lets remove the var for clarity!!
```


# Case: Personal net hourly wage by sex and education 

Lets then do a quick demo on how to use this package with a simple analysis that uses following variables:

- *pmi* = `pmi`
- *gender* = `sex`
- *highest completed education level* = `educ`

Lets print the metadata out of those variables

```{r}
# re-creating metadata
spss  <- read_sav("./datasets/data.sav")
spss_meta <- create_metadata(spss)
meta_spss[meta_spss$code %in% c("sex","pmi","educ"),] %>% 
  kable()
```

**Note**

*In this particular data they have added numeric prefixes for each value label to define the order of the levels. As it looks odd in the outputs, lets remove all with some reqular expression*

```{r}
meta_spss$label <- sub("^\\[[0-9]+\\]", "", meta_spss$label)
#
meta_spss[meta_spss$code %in% c("sex","pmi","educ"),] %>% 
  kable()
```


To get started with the analysis, lets first compute the mean hourly wage for each category using the stripped data.

```{r}
data_spss %>% 
  select(cname,sex,educ,pmi) %>% 
  na.omit() %>%
  group_by(cname,sex,educ) %>% 
  summarise(mean_wage = mean(pmi)) -> d_mean

library(ggplot2)
ggplot(d_mean, aes(x=educ,y=mean_wage)) + geom_col() + facet_wrap(~sex)
```

Then lets label the variables before plotting it with `into.factor=FALSE`

```{r}
data_spss %>% 
  select(cname,sex,educ,pmi) %>% 
  na.omit() %>%
  group_by(cname,sex,educ) %>% 
  summarise(mean_wage = mean(pmi)) %>% 
  ungroup() %>% 
  mutate(sex = label_data(.,variable.data = "sex", metadata = meta_spss, into.factor=FALSE),
         educ = label_data(.,variable.data = "educ", metadata = meta_spss, into.factor=FALSE)) -> d_mean

ggplot(d_mean, aes(x=educ,y=mean_wage)) + geom_col() + facet_wrap(~sex)
```

With `into.factor` set as `FALSE` the value labels are put in place as character ie. without any order. When set `TRUE` the value labels are forced into factors with same order of value labels as in the original SPSS/Stata files. 

```{r}
data_spss %>% 
  select(cname,sex,educ,pmi) %>% 
  na.omit() %>%
  group_by(cname,sex,educ) %>% 
  summarise(mean_wage = mean(pmi)) %>% 
  ungroup() %>% 
  mutate(sex = label_data(.,variable.data = "sex", metadata = meta_spss, into.factor=TRUE),
         educ = label_data(.,variable.data = "educ", metadata = meta_spss, into.factor=TRUE)) -> d_mean

ggplot(d_mean, aes(x=educ,y=mean_wage)) + geom_col() + facet_wrap(~sex)
```

Package [`forcats`](http://forcats.tidyverse.org/) is useful when you need to do more adjusting for the factors, for instance reversing the order of the factor labels as below.

```{r}
data_spss %>% 
  select(cname,sex,educ,pmi) %>% 
  na.omit() %>%
  group_by(cname,sex,educ) %>% 
  summarise(mean_wage = mean(pmi)) %>% 
  ungroup() %>% 
  mutate(sex = label_data(.,variable.data = "sex", 
                          metadata = meta_spss, into.factor=TRUE),
         educ = label_data(.,variable.data = "educ", 
                           metadata = meta_spss, into.factor=TRUE)) %>% 
  mutate(
    sex  = forcats::fct_rev(sex),
    educ = forcats::fct_rev(educ)
  ) -> d_mean

ggplot(d_mean, aes(x=educ,y=mean_wage)) + geom_col() + facet_wrap(~sex)
```

## Adding new variables

You often need to create new variables when working with social surveys, this is how you would do for `age_class`

First, lets make sure that variable ´age` is numeric with

```{r}
meta_spss[meta_spss$code == "age",]$class == "numeric"
```

then classify it into four categories

```{r}
data_spss$age_class[data_spss$age < 20] <- 1
data_spss$age_class[data_spss$age >= 20 & data_spss$age < 40] <- 2
data_spss$age_class[data_spss$age >= 40 & data_spss$age < 60] <- 3
data_spss$age_class[data_spss$age >= 60] <- 4
as.data.frame(table(data_spss$age_class))
```

then lets add the new variable into metadata

```{r}
new_row <- data_frame(
  code = "age_class",
  name = "age in four classes",
  value = 1:4,
  label = c("under 25",
            "25 to 39",
            "40 to 59",
            "60 or more"),
  class="factor"
)
meta_spss <- bind_rows(meta_spss,new_row)
```

Lets add the age class dimension to our analysis from above

```{r}
data_spss %>% 
  select(cname,sex,educ,pmi,age_class) %>% 
  na.omit() %>%
  group_by(cname,sex,educ,age_class) %>% 
  summarise(mean_wage = mean(pmi)) %>% 
  ungroup() %>% 
  mutate(sex = label_data(.,variable.data = "sex", metadata = meta_spss, into.factor=TRUE),
         educ = label_data(.,variable.data = "educ", metadata = meta_spss, into.factor=TRUE),
         age_class = label_data(.,variable.data = "age_class", metadata = meta_spss, into.factor=TRUE)) -> d_mean

ggplot(d_mean, aes(x=educ,y=mean_wage)) + geom_col() + facet_grid(age_class~sex)
```

Again, this approach is very useful when working with shiny-applications as pointed out in this example at <https://github.com/muuankarski/tidymetadata/tree/master/inst/shiny-examples/lisapp> that can be lauched with `tidymetadata::runExample()`


# Re-create the spss-data

You can then re-create spss-data from stripped data & meta-data using the following script

**Add the variable labels ie. variable names or questions**

```{r}
d <- data_spss
meta <- meta_spss

library(labelled)
for (n in names(d)){
  var_label(d[[n]]) <- meta[meta$code %in% n, ]$name[1]
}
```

**Add the value labels ie. 1 equals male, 2, equals female etc.**

```{r}
for (n in names(d)){
  # if numeric, no need for value label
  if (meta[meta$code %in% n,]$class[1] %in% c("numeric","character")) next()
  vec <- as.integer(meta[meta$code %in% n, ]$value)
  names(vec) <- meta[meta$code %in% n, ]$label
  d[[n]] <- labelled(d[[n]], label=vec)
}
```

And this is how the first 15 variables look like with the attributes

```{r}
str(d[1:15])
```

And you can then save it as .sav using haven

```{r, eval=FALSE}
haven:write_sav(d, path = "./lis_spss.sav")
```

